Thesis: draft

### 2021-03-27 draft.tex
made some small adjustments and continued writing in chapter 4

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\ Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are discussed.

@@

Some other important concepts in network theory are the degree distribution, clustering/community structure, connectivity, etc. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate a bit on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering. These two are also referred to as "small world phenomenon" (note that in the literature the term "small world" sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of these three properties. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of node $i$. The degree can be easily obtained from the adjacency matrix \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\text{in}}$ and the number of outgoing edges $k_i^{\text{out}}$ of node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{equation}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \quad \text{and} \quad k_i^{\text{out}} = \sum_{j=1}^N A_{ij}
\end{equation}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij}
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represents the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k)
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution also allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree close to the average degree $\left<k\right>$. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. % explain differences  more, consequences of power-law + need to include figure?
They are also referred to as scale free networks, since they do not posses a characteristic length scale or, in this case, the average degree isn't a characteristic scale for the network \cite{Mata2020}. % correct statement? + talk about fact that av degree can diverge; explain scale free concept a bit more detailed
%maybe explain why we care about degree, centrality measure etc

@@

There are no real restrictions to the edge probabilities, which makes the stochastic block model a very general model that is capable of reproducing many different network structures \cite{Karrer2011}. However, the edge probabilities inside a community are often taken to be bigger than the probabilities between communities ($M_{ii} > M_{ij}$, $i \neq j$). In this way, the stochastic block model generates a network with a clear community structure. A stochastic block model that fulfills this condition is called an assortative SBM \cite{Lee2019}. A disassortative stochastic block model refers to the case where the edge probabilities between the communities are higher than those inside the communities ($M_{ii} < M_{ij}$, $i \neq j$) \cite{Gribel2020}.\\

@@

The stochastic block model includes the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. This can be seen in the two different ways. First, consider the case where $k = 1$, thus we only have one community and all nodes belong to this community. The matrix $M$ then reduces to one number $p$ and we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. Second, assume we have more than one group or community, $k > 1$, and assume that every element of the matrix $M$ is the same, $M_{ij} = p$ and $M_{ii}=p$ \cite{Clauset2017}. Again, we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model. In both cases, the mathematical properties of the Erd\H{o}s-R\'{e}nyi model are obtained: a Poisson degree distribution, an average distance that is logarithmic with the system size $N$, vanishing clustering coefficient for large $N$, etc \cite{Clauset2017}. The structure within the groups or communities of the stochastic block model is that of a random graph, while the structure between the communities is that of a random bipartite graph \cite{Clauset2017}.\\

@@

% talk about extention with WS as constituents instead of ER

@@

where the exponent $\gamma$ often lies between 2 and 3. Networks with a power-law degree distribution are also called heterogeneous or scale free networks. The term scale free refers to the fact that there is no characteristic scale in the network. Compared to the homogeneous network, where each node has a degree close to the average degree $\left<k\right>$, the heterogeneous network has many nodes with a low degree and a few nodes with a very high degree (these nodes are referred to as hubs). If the exponent $\gamma$ is smaller or equal to two ($\gamma \leqslant 2$), the average degree diverges; if $2 < \gamma < 3$, there is a finite average degree, but a diverging variance \cite{Newman2005}.\\

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fractions of neighbors of user $i$ with opinion A and B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will update his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

Another extremely simple binary model, maybe even simpler than the majority model, is the voter model. The model works as follows: each agent in the network has a binary opinion (A or B); at each time step, an agent is randomly selected along with one of its neighbors and, subsequently, the agent adopts the opinion of this neighbor \cite{Castellano2009}. The agent thus mimics its neighbor \cite{Castellano2009}. \\
The pressure of the majority in only felt in an averaged way, in the sense that a neighbor that carries the majority opinion has more chance to be randomly selected \cite{Castellano2009}.\\
Lots of studies have been performed on this model. In this thesis, however, we will not make use of the voter model. Hence, we will not go into further detail. % cite papers that have done studies on this model??
% https://arxiv.org/pdf/0710.3256.pdf

@@

In the classical model, there are $N$ nodes/agents, each with an opinion $x_i$ and each node $i$ gives a weight $w_{ij}$ to all its neighbors (with $w_{ij} \geqslant 0$; this weight may for example represent the importance/influence of the neighbors) \cite{Krause2002}. The most simple model has fixed weights $w$ and the updating rule is \cite{Krause2002}
\begin{equation}\label{refMod}
	x(t+1) = W x(t)
\end{equation}
where $W$ is the fixed stochastic matrix which contains the weights $w_{ij}$, $x(t)$ is the opinion profile at time $t$ (this is a $N$-dimensional vector containing the opinions $x_i$ of each node at time $t$) and $x(t+1)$ is the opinion profile at time $t+1$.\\
Friedkin and Johnsen proposed a variation of this model \cite{Friedkin1990}\cite{Friedkin1999}. The model they proposed assumes that an agent $i$ adheres to its initial opinion to a certain degree $g_i$ and is influenced by others by a susceptibility of $1-g_i$ \cite{Krause2002}. Thus this variation has the following updating rule (in matrix notation) \cite{Krause2002}
\begin{equation}
	x(t+1) = Gx(0) + (\mathbb{1}-G)Wx(t)
\end{equation}
where $G$ is the diagonal matrix which contains the $g_i$ and $\mathbb{1}$ is the identity matrix. If all the $g_i$ are equal to zero, we obtain the model of Eq. (\ref{refMod}) \cite{Krause2002}.\\
A possible time variant extension is \cite{Krause2002}
\begin{equation}
	x(t+1) = W(t)x(t)
\end{equation}
where now the weights are time dependent. This model can be used for situations in which so called "hardening of positions" occur \cite{Krause2002}. This means that, over time, the agents put more and more weight on their own opinion and less weight on the opinion of others \cite{Krause2002}. 
% talk about extension http://jasss.soc.surrey.ac.uk/5/3/2/2.pdf

@@

In bounded confidence models, each node has, beside an opinion, a threshold $\epsilon$, also called an uncertainty or tolerance (note that $\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. The two most popular bounded confidence models are the Deffuant model and the Hegselmann-Krause model \cite{Castellano2009}. \\

@@

The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (so the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinions inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends in the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{"Modelling opinion dynamics in the age of algorithmic personalization"} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $Q$ elements. This filtering method is based on the semantic filtering explained above
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC and REF methods are mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

@@

If a user is active, he/she first spreads his current opinion to all his/her neighbors (for example by posting something). Then, the active user updates his own opinion according to the opinions on his/her time line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.\\
It is important to note that what somebody sees on his/her time line is the consequence of the interplay between the filtering algorithm and the posted opinions of his/her neighbors. Every user has a "hidden time line $L_i(t)$" which contains the actual time line of the user at the moment when he/she was active for the last time ($t_{\text{last}}$) plus all the posts of the neighbors from that last time he/she was active till the current time he/she becomes active again ($t$) \cite{Perra2019}. At the time $t$ that the user becomes active again, the filtering algorithm selects $Q$ post from this "hidden time line $L_i(t)$" and shows them on the actual time line of the user; these $Q$ items are thus the opinions that the user is exposed to \cite{Perra2019} (see also Section \ref{filter} for more detailed information on the filtering algorithms).

@@

\section{The presence of stubborn people}\label{stubb}

>>>>

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time-line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\ Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are discussed.

@@

Some other important concepts in network theory are the degree distribution, clustering/community structure, connectivity, etc. These concepts will be explained in depth in Chapter \ref{chap2}, Section \ref{netDef}, but let us already anticipate a bit on the case of social networks. It is found that many real-life social systems have a power-law degree distribution \cite{Muchnik2013}. This heterogeneous or scale-free degree distribution represents one of the three general properties of social networks. The other two are short distances and high clustering. These two are also referred to as "small world phenomenon" (note that in the literature the term "small world" sometimes only refers to the property of short distances, however, in this thesis, we will use it to indicate both a high clustering and short distances) \cite{Muchnik2013}. Ideally, our theoretical network should exhibit these three properties. Some theoretical networks, that are thoroughly studied, do not possess all of these three properties. However, they might still be used, because of their simplicity and ability to produce analytic results. When using these models one must always bear in mind their limitations to reproduce some properties encountered in real social systems. \\

@@

The adjacency matrix contains a lot of information such as the degree of a node. The degree $k_i$ of a node $i$ is the number of edges attached to node $i$ or, in others words, $k_i$ represents the number of nearest neighbors of node $i$. The degree can be easily obtained from the adjacency matrix \cite{Mata2020}
\begin{equation}
	k_i = \sum_{j=1}^N A_{ij}
\end{equation}
In the case of a directed graph one can differentiate between the number of incoming edges $k_i^{\text{in}}$ and the number of outgoing edges $k_i^{\text{out}}$ of node $i$. These are called the in-degree and out-degree respectively and are defined as \cite{Mata2020}
\begin{equation}
	k_i^{\text{in}} = \sum_{j=1}^N A_{ji} \quad \text{and} \quad k_i^{\text{out}} = \sum_{j=1}^N A_{ij}
\end{equation}
The total degree of node $i$ is then $k_i = k_i^{\text{in}} + k_i^{\text{out}}$. For weighted graphs the degree is easily generalized to the weighted degree $s_i$, often called strength \cite{Ioannis2007}
\begin{equation}
	s_i = \sum_{j=1}^N w_{ij}
\end{equation}
From the degree of the nodes in the network one can construct the degree distribution. The degree distribution $P(k)$ represents the fraction of nodes with a degree $k$ or, in other words, the degree distribution gives the probability that a randomly selected node has a degree $k$. The average degree can be obtained from the degree distribution \cite{Mata2020}
\begin{equation}
	\left<k\right> = \frac{1}{N} \sum_{i=1}^N k_i = \sum_k k P(k)
\end{equation}
where $N$ is the total number of nodes in the network. The degree distribution also allows us to classify networks. The two most important classes are homogeneous and heterogeneous networks. Homogeneous networks have a bell curved degree distribution, eg. a Poisson distribution. In this case, most of the nodes have a degree close to the average degree $\left<k\right>$. Heterogeneous networks, on the other hand, have a power-law degree distribution, $P(k) \sim k^{-\gamma}$. % explain differences  more, consequences of power-law + need to include figure?
They are also referred to as scale-free networks, since they do not posses a characteristic length scale or, in this case, the average degree isn't a characteristic scale for the network \cite{Mata2020}. % correct statement? + talk about fact that av degree can diverge; explain scale free concept a bit more detailed
%maybe explain why we care about degree, centrality measure etc

@@

There are no real restrictions on the edge probabilities, which makes the stochastic block model a very general model that is capable of reproducing many different network structures \cite{Karrer2011}. However, the edge probabilities inside a community are often taken to be bigger than the probabilities between communities ($M_{ii} > M_{ij}$, $i \neq j$). In this way, the stochastic block model generates a network with a clear community structure. A stochastic block model that fulfills this condition is called an assortative SBM \cite{Lee2019}. A disassortative stochastic block model refers to the case where the edge probabilities between the communities are higher than those inside the communities ($M_{ii} < M_{ij}$, $i \neq j$) \cite{Gribel2020}.\\

@@

The stochastic block model includes the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. This can be seen in the two different ways. First, consider the case where $k = 1$, thus we only have one community and all nodes belong to this community. The matrix $M$ then reduces to one number $p$ and we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model \cite{Clauset2017}. Second, assume we have more than one group or community, $k > 1$, and assume that every element of the matrix $M$ is the same, $M_{ij} = p$, $\forall \ i, j$ \cite{Clauset2017}. Again, we are left with the parameters $N$ and $p$ of the Erd\H{o}s-R\'{e}nyi model. In both cases, the mathematical properties of the Erd\H{o}s-R\'{e}nyi model are obtained: a Poisson degree distribution, an average distance that is logarithmic with the system size $N$, vanishing clustering coefficient for large $N$, etc \cite{Clauset2017}. The structure within the groups or communities of the stochastic block model is that of a random graph, while the structure between the communities is that of a random bipartite graph \cite{Clauset2017}.\\

@@

@@

where the exponent $\gamma$ often lies between 2 and 3. Networks with a power-law degree distribution are also called heterogeneous or scale-free networks. The term scale-free refers to the fact that there is no characteristic scale in the network. Compared to the homogeneous network, where each node has a degree close to the average degree $\left<k\right>$, the heterogeneous network has many nodes with a low degree and a few nodes with a very high degree (these nodes are referred to as hubs). If the exponent $\gamma$ is smaller or equal to two ($\gamma \leqslant 2$), the average degree diverges; if $2 < \gamma < 3$, there is a finite average degree, but a diverging variance \cite{Newman2005}.\\

@@

The probabilistic majority model is similar to the majority model, but has, as the name suggests, a more probabilistic/random nature. In this model the fraction of neighbors of user $i$ with opinion A and the fraction of neighbors with opinion B are determined. Then, the opinion of user $i$ is updated: he/she takes on opinion A (or B) with a probability equal to the fraction of his/her neighbors with opinion A (or B) \cite{Perra2019}.\\
For example, if $70\%$ of the neighbors of user $i$ have an opinion A (and thus $30 \%$ of the neighbors have opinion B), user $i$ will update his/her opinion to opinion A with a probability 0.7 and to opinion B with a probability 0.3 \cite{Perra2019}. Thus, user $i$ no longer simply updates his/her opinion according to the majority of opinions carried by his/her neighbors, but the majority opinion still has a higher probability to be taken on by this user.\\
This model is no longer deterministic, but has a level of randomness, which is an important component of opinion dynamics \cite{Perra2019}. 
% https://arxiv.org/pdf/0710.3256.pdf read this (overview, discusses several models)!!

@@

Another extremely simple binary model, maybe even simpler than the majority model, is the voter model. The model works as follows: each agent in the network has a binary opinion (A or B); at each time step, an agent is randomly selected along with one of its neighbors and, subsequently, the agent adopts the opinion of this neighbor \cite{Castellano2009}. The agent thus mimics its neighbor \cite{Castellano2009}. \\
The pressure of the majority in only felt in an averaged way, in the sense that a neighbor that carries the majority opinion has more chance to be randomly selected \cite{Castellano2009}.\\
Lots of studies have been performed on this model. In this thesis, however, we will not make use of the voter model. Hence, we will not go into further detail. Interested readers are referred to references \cite{Castellano2009}, \cite{Fernandez2014} and related references therein.% cite papers that have done studies on this model??

@@

In the classical model, there are $N$ nodes/agents, each with an opinion $x_i$ and each node $i$ gives a weight $w_{ij}$ to all its neighbors (with $w_{ij} \geqslant 0$; this weight may for example represent the importance/influence of the neighbors) \cite{Krause2002}. The most simple model has fixed weights $w$ and the updating rule is \cite{Krause2002}
\begin{equation}\label{refMod}
	x(t+1) = W x(t)
\end{equation}
where $W$ is the fixed stochastic matrix which contains the weights $w_{ij}$, $x(t)$ is the opinion profile at time $t$ (this is a $N$-dimensional vector containing the opinions $x_i$ of each node at time $t$) and $x(t+1)$ is the opinion profile at time $t+1$.\\
Friedkin and Johnsen proposed a variation of this model \cite{Friedkin1990}\cite{Friedkin1999}. The model they proposed assumes that an agent $i$ adheres to its initial opinion to a certain degree $g_i$ and is influenced by others by a susceptibility of $1-g_i$ \cite{Krause2002}. This variation thus has the following updating rule (in matrix notation) \cite{Krause2002}
\begin{equation}
	x(t+1) = Gx(0) + (\mathbb{1}-G)Wx(t)
\end{equation}
where $G$ is the diagonal matrix which contains the $g_i$ and $\mathbb{1}$ is the identity matrix. If all the $g_i$ are equal to zero, we obtain the model of Eq. (\ref{refMod}) \cite{Krause2002}.\\
A possible time variant extension is \cite{Krause2002}
\begin{equation}
	x(t+1) = W(t)x(t)
\end{equation}
where now the weights are time dependent. This model can be used for situations in which so called "hardening of positions" occur \cite{Krause2002}. This means that, over time, the agents put more and more weight on their own opinion and less weight on the opinion of others \cite{Krause2002}. 
% talk about extension http://jasss.soc.surrey.ac.uk/5/3/2/2.pdf

@@

In bounded confidence models, each node has, beside an opinion $x_i$, a threshold $\epsilon_i$, also called an uncertainty or tolerance (note that $\epsilon$ can be the same for all nodes or can be different for each node) \cite{Castellano2009}\cite{Deffuant2006}. This threshold determines which opinions of neighbors are close enough to the agent's opinion in order to influence that agent; opinions that are beyond this threshold are ignored \cite{Deffuant2006}. In other words, an agent with opinion $x_i$ is only affected/influenced by those neighbors which have an opinion in the interval $\interval{x_i-\epsilon}{x_i+\epsilon}$ \cite{Castellano2009}. \\
The two most popular bounded confidence models are the Deffuant model and the Hegselmann-Krause model \cite{Castellano2009}. \\

@@

The final state is a situation where several opinion clusters coexist. This is because the opinions move closer and closer to each other in several clusters; once each cluster is sufficiently far apart from the others (so the difference in opinions between all agents in the two clusters is larger than $\epsilon$), they no longer interact with each other \cite{Castellano2009}. Only agents inside the same cluster keep on interacting with each other, which makes the opinions inside a cluster converge to a single value \cite{Castellano2009}. Usually, the number and sizes of the clusters depends on the threshold $\epsilon$, whereas $\mu$ determines the convergence rate \cite{Castellano2009}.\\

@@

On-line social platforms use algorithms to order and filter the posts that appear on the time-line of an individual \cite{Perra2019}. They do this in order to make the experience on their platforms more pleasant, interesting and convenient \cite{Perra2019}. If social media companies woudn't use any kind of filtering, there would be way to many content to handle and process for the social media user \cite{Bozdag2013}. This in turn would make the on-line social experience a lot more unpleasant and inconvenient; the user might feel overwhelmed by the number of information he/she is exposed to \cite{Bozdag2013}. The reason behind this is that people are cognitive and time constrained \cite{Perra2019}; hence, there is only a limited amount of information that humans can process \cite{Bozdag2013}.\\
\newline
The backside of such algorithmic filtering is that it fine-tunes the information it shows based on the user. Even two users with the same friends may then be exposed to different content, based on their previous activities and history \cite{Bozdag2013}. This may lead to the formation of echo chambers in which users get `trapped' in content that confirms their own mindset \cite{Bozdag2013}. \\
It is thus important to know and be aware of the possible impacts of algorithmic personalization on the formation and evolution of opinions. Prof. Nicola Perra and prof. Luis E.C. Rocha already partly investigated the role of such algorithms on the dynamics of opinions in several networks in one of their previous papers, \textit{"Modelling opinion dynamics in the age of algorithmic personalization"} (see reference \cite{Perra2019}). Since filtering algorithms are often not transparently shared by social media companies, they proposed several personalization algorithms based on three simple ways of filtering \cite{Perra2019}
\begin{itemize}
	\item Popularity filtering: promoting content that is popular across the platform
	\item Semantic filtering: promoting posts that are similar to previous consumed posts
	\item Collaborative filtering: promoting content that people similar or close to us consume
\end{itemize} 
The following paragraph briefly reviews the filtering algorithms that prof. Nicola Perra and prof. Luis E.C. Rocha proposed. \\
\newline
At every time $t$ each platform user has a hidden list $L_i(t)$ that contains all the opinions/posts that he/she received from his/her friends \cite{Perra2019}. The filtering algorithm then selects $Q$ items from this list and shows them on the user's time-line (this is the fundamental idea behind algorithmic personalization: only showing a fraction of the possible posts \cite{Perra2019}); the following selection methods are used \cite{Perra2019}
\begin{itemize}
	\item The reference method (REF): this method randomly selects $Q$ items from the list $L_i(t)$
	\item The old method (OLD): the $Q$ oldest post in $L_i(t)$ are displayed on the time-line of the user
	\item The recent method (REC): shows the $Q$ most recent items from $L_i(t)$
	\item The preference method (PR): sorts the hidden list $L_i(t)$ according to the current opinion of the user and then selects the first $Q$ elements. This filtering method is based on the semantic filtering explained above
\end{itemize}
Prof. Nicola Perra and prof. Luis E.C. Rocha showed that the REF, OLD and REC methods behave similar in many cases and that the PR method often displays different behavior (such as a larger formation of echo chambers compared to the other three methods) \cite{Perra2019}. In this thesis the REC and REF methods are mainly used as a reference/benchmark method and the PR method is used as the main filtering algorithm.

@@

If a user is active, he/she first spreads his current opinion to all his/her neighbors (for example by posting something). Then, the active user updates his own opinion according to the opinions on his/her time-line \cite{Perra2019}. This updating goes according to the implemented opinion dynamics model (eg. majority model, probabilistic majority model,...). In every time step the updating of all active nodes is done simultaneously \cite{Perra2019}.\\
It is important to note that what somebody sees on his/her time-line is the consequence of the interplay between the filtering algorithm and the posted opinions of his/her neighbors. Every user has a "hidden time-line $L_i(t)$" which contains the actual time-line of the user at the moment when he/she was active for the last time ($t_{\text{last}}$) plus all the posts of the neighbors from that last time he/she was active till the current time he/she becomes active again ($t$) \cite{Perra2019}. At the time $t$ that the user becomes active again, the filtering algorithm selects $Q$ post from this "hidden time-line $L_i(t)$" and shows them on the actual time-line of the user; these $Q$ items are thus the opinions that the user is exposed to \cite{Perra2019} (see also Section \ref{filter} for more detailed information on the filtering algorithms).

@@

\newpage
\section{The presence of stubborn people}\label{stubb}

It is well known that in real life, you can find people that are really persistent towards their own opinion and that are not open to opinions of others. They are convinced that they carry the truth with them and that other people, with different world-views, are not capable of giving them any new and useful insights. They are close-minded in contrast to people with an open-mind that are willing to learn from others and to reconsider their beliefs. Of course this is put a bit simplistic and there are many shades of gray between this black and white statement. However, there are groups in society, often with more extreme opinions, that are know to be really inflexible towards changing that (extreme) opinion (note here that being stubborn and being extreme doesn't need to go hand in hand, you can obviously be stubborn without being extreme and vice versa; however, it is often widely accepted that there is some kind of correlation/link between people with extreme opinions and being more persistent towards these opinions compared to people with more moderate beliefs; this idea of correlating extreme people with a more stubborn character is also often used in models, see for example \cite{Deffuant2006}\cite{Turner}) (also note that having some resistance to change your opinion to any other opinion doesn't necessarily need to be a bad thing, it can be seen as a form of critical thinking and as a form of taking things with a grain of salt).\\ 
\newline
Knowing the impact of such stubborn groups on the opinions of others is of great importance, especially with the rise of social media which makes the reach of such extreme and persistent groups a lot bigger. Before the era of social media, one could maybe argue that these extreme and stubborn groups are often isolated and therefore wouldn't reach too many people. So their impact could only be of a minor nature (note that this is only a statement that could have been given, there are of course plenty of examples throughout human history that contradict this argument; to only mention a few: Nazism in the 1930s and 1940s, the Ku Klux Klan in the late 19th and begin 20th century, the rise of communism in the 20th century, etc, all examples of groups that can be considered extreme and that have had a wide reach and a significant impact on society). Today, this possible argument is no longer valid; social media enables the contact with all sorts of beliefs and all sorts of people with almost no (time and spatial) constraints. Extreme and stubborn groups that may have been isolated before no longer need to be so and this may have huge, and largely unknown, impacts. \\
\newline
What is the impact of having stubborn/extreme people in your society? Does it matter if the stubborn people are grouped together or if they are distributed evenly throughout your system? Are there ways to reduce the possible impacts of stubbornness and/or extremism? Can these stubborn groups lead to polarization or even overtake the opinions in the system? Or is introducing some kind of stubbornness, on the other hand, a protective factor towards extremism? To answer these questions, several models have been proposed to include some kind of stubbornness/resistance in the system under study.
